---
title: "Linear Regressions"
author: "dnkimani"
date: "2024-11-28"
output: html_document
---

# Linear Regression
## 1.1 Setting up the Data
```{r}
#library a bunch of packages we may (or may not) use - install them first if not installed already. 
library(tidyverse)
library(tmap)
library(plotly)
library(broom)
library(mapview)
library(sf)
library(sp)
library(spdep)
library(car)
library(fs)
library(janitor)
```
Lets read in some ward data.
```{r}
#download a zip file containing some boundaries we want to use

download.file("https://data.london.gov.uk/download/statistical-gis-boundary-files-london/9ba8c833-6370-4b11-abdc-314aa020d5e0/statistical-gis-boundaries-london.zip", 
              destfile="Prac8_data/statistical-gis-boundaries-london.zip")
```

Get the zip file and extract it
```{r}
#library(fs)
listfiles<-dir_info(here::here("Prac8_data")) %>%
  dplyr::filter(str_detect(path, ".zip")) %>%
  dplyr::select(path)%>%
  pull()%>%
  #print out the .gz file
  print()%>%
  as.character()%>%
  utils::unzip(exdir=here::here("Prac8_data"))
```

Look inside the zip and read in the `.shp`
```{r}
#look what is inside the zip

Londonwards<-fs::dir_info(here::here("Prac8_data", 
                                 "statistical-gis-boundaries-london", 
                                 "ESRI"))%>%
  #$ means exact match
  dplyr::filter(str_detect(path, 
                           "London_Ward_CityMerged.shp$"))%>%
  dplyr::select(path)%>%
  dplyr::pull()%>%
  #read in the file in
  sf::st_read()
```

A quick check on the data.
```{r}
#check the data
qtm(Londonwards)
```

Now we read in some additional data from the [London Datastore]("https://data.london.gov.uk/")

```{r}
#read in some attribute data
LondonWardProfiles <- read_csv("https://data.london.gov.uk/download/ward-profiles-and-atlas/772d2d64-e8c6-46cb-86f9-e52b4c7851bc/ward-profiles-excel-version.csv", 
                               col_names = TRUE, 
                               locale = locale(encoding = 'Latin1'))

#check all of the columns have been read in correctly
Datatypelist <- LondonWardProfiles %>% 
  summarise_all(class) %>%
  pivot_longer(everything(), 
               names_to="All_variables", 
               values_to="Variable_class")

Datatypelist
```

### 1.1.1 Cleaning the Data
```{r}
#We can use readr to deal with the issues in this dataset - which are to do with text values being stored in columns containing numeric values

#read in some data - couple of things here. Read in specifying a load of likely 'n/a' values, also specify Latin1 as encoding as there is a pound sign (£) in one of the column headers - just to make things fun!

LondonWardProfiles <- read_csv("https://data.london.gov.uk/download/ward-profiles-and-atlas/772d2d64-e8c6-46cb-86f9-e52b4c7851bc/ward-profiles-excel-version.csv", 
                               na = c("", "NA", "n/a"), 
                               locale = locale(encoding = 'Latin1'), 
                               col_names = TRUE)

#check all of the columns have been read in correctly
Datatypelist <- LondonWardProfiles %>% 
  summarise_all(class) %>%
  pivot_longer(everything(), 
               names_to="All_variables", 
               values_to="Variable_class")

Datatypelist
```

```{r eval=FALSE, include=FALSE}
# This won't run, it's just for an alternative method
#you could download it from the London data store and read it in…it’s the ward Profiles excel download
LondonWardProfiles <- read_csv("prac7_data/ward-profiles-excel-version.csv", 
                               na = c("", "NA", "n/a"), 
                               locale = locale(encoding = 'Latin1'), 
                               col_names = TRUE)
```

Now we will merge `Londonwards` with `LondonWardProfiles` using the appropriate codes
```{r}
#merge boundaries and data
LonWardProfiles <- Londonwards%>%
  left_join(.,
            LondonWardProfiles, 
            by = c("GSS_CODE" = "New code"))

#let's map our dependent variable to see if the join has worked:
tmap_mode("plot")
qtm(LonWardProfiles, 
    fill = "Average GCSE capped point scores - 2014", 
    borders = NULL,  
    fill.palette = "Blues")
```

### 1.1.2 Additional Data
```{r}
#might be a good idea to see where the secondary schools are in London too
london_schools <- read_csv("https://data.london.gov.uk/download/london-schools-atlas/57046151-39a0-45d9-8dc0-27ea7fd02de8/all_schools_xy_2016.csv")

#from the coordinate values stored in the x and y columns, which look like they are latitude and longitude values, create a new points dataset
lon_schools_sf <- st_as_sf(london_schools, 
                           coords = c("x","y"), 
                           crs = 4326)

lond_sec_schools_sf <- lon_schools_sf %>%
  filter(PHASE=="Secondary")

tmap_mode("plot")
qtm(lond_sec_schools_sf)
```

## 2.Analysing GCSE exam performance - testing a research hypothesis 